---
title: "Lab 8"
output: pdf_document
---

Your next deliverable is on Data Visualizations. This document will give some 
examples of plots that can be done using `ggplot2`. You will need to import the
`ggplot2` library to work with the functions.

```{r setup, include=TRUE}
library(ggplot2)
# We will give some examples 
data("diamonds")
```

This lab will give some examples of plots to be included in your paper.

## Bar plots

You will want to check that your population is representative. One way to do that
is to see if the proportion of samples landing in certain classes is what you
expect. 

Here we look at the counts of diamonds of different colors in our data set.

``` {r, echo=TRUE}

BarPlot <- ggplot(data = diamonds, aes(x = color, fill = color)) + 
  geom_bar(stat = "count") + 
  ggtitle("Color of diamonds")
print(BarPlot)
```

If we expected a even distribution of colors, then we can tell from this graph that
our sample is likely not representative. For instance, the number of color "J" 
diamonds is less than one-third of color "G" diamonds. The distribution into classes
is dependent on the context of your problem. If we know ahead of time that color "J"
is naturally rarer, then this sample would not necessarily be un-representative.

## Scatterplot

When comparing two quantitative variables, you may be testing whether $\beta_1 \neq 0$.
For this test, you will need to assume that the variables are linearly related.
You can check this visually by creating a scatter plot with the line of best
fit graphed.

Here we compare the Price of diamonds with the carats.

```{r}
ScatterPlot <- ggplot(data = diamonds, aes(x=carat,y=price)) + geom_point() +
  geom_smooth(method = "lm", se = FALSE) + ggtitle("Price vs Carat")
print(ScatterPlot)
```
From this, we can see that the assumption of the linear relationship is mostly 
violated. For higher carat values, the expected price is not described by the line.
There seems to be some sort of (perhaps artificial?) maximum price of $20,000.

However, we can still see that there is a strong relationship between carats and
price (especially between 0 and 1). For the purposes of the project, we may decide
to continue with the hypothesis test despite our assumption being violated. In some
cases it may be more appropriate to transform the data to get a linear relationship.
A common transformation would be to take the logarithm of one or both of the variables:

```{r}
ScatterPlot <- ggplot(data = diamonds, aes(x=log(carat),y=log(price))) + geom_point() +
  geom_smooth(method = "lm", se = FALSE) + ggtitle("log(Price) vs log(Carat)")
print(ScatterPlot)
```
After taking the natural log of both variables, then the linear relationship looks
much stronger! So we may want to change our test to assume a linear relationship
between the logarithms of both variables.

# Comparing populations

When comparing means from different populations (like trying to show $\mu_1> \mu_2$ for example)
it is useful to look at the distributions of each population side-by-side. Here
we are interested in comparing the price of the diamonds based on the cut. One 
way to compare distributions is with a violin plot.

```{r}
ViolinPlot <- ggplot(data = diamonds, aes(x=cut, y = price, fill=cut)) + 
  geom_violin() +
  ggtitle("Price vs Cut of Diamonds")
print(ViolinPlot)
```
Here we can see that all the distributions are right-skewed (there are a few diamonds
worth a lot of money for each cut). Comparing the Ideal and Fair cuts, the 
Ideal cut seem to have the prices strongly centered around $1000 while the Fair
cut is more evenly distributed from \$0 to $5000. This suggests that the Fair
cuts diamonds are more valuable on average than the Ideal cut.

If you are comparing proportions from different populations, then a stacked bar
plot can be useful for that. Here we look at the percentage of colors for each cut
of diamond:
```{r}
StackedBar <- ggplot(diamonds,aes(cut,fill=color))+geom_bar(position="fill") +
  ggtitle("Percentage of colors based on cut") + ylab("Percentage")
print(StackedBar)
```

Looking at the graph, we don't notice too much of a difference between the cuts.
Depending on what we are interested in, we can make further observations. If we 
were interested in the proportion of color "E", the we could note that the Fair cut
seems to have a smaller percentage of that color than other cuts.