---
title: 'Lab 6: Data Visualizations with `ggplot2`'
output: pdf_document
---

```{r setup, include=FALSE}
library(ggplot2)
```

## Introduction to `ggplot2`

Previously, we have done plotting and graphs with the built-in function for R.
However, in this lab we will start using `ggplot2` to create attractive visuals.
This library will be useful for doing the data visualization for your project.

**Task**: Check if `ggplot2` is installed on your system. You can do this by 
simply running `?ggplot2` in your console and seeing if there is documentation.
If you need to install this library, run the following
``` 
install.packages("ggplot2")
```

For this lab, we will be saving and building plots using `ggplot2`, so I would
suggest creating a new R script to document your work. To import the library 
into your current environment, run `library(ggplot2)`.

The `ggplot2` library is based on the "grammar of graphics", the idea being that
you can build every graph from the same components: a data set, a coordinate
system, and geoms (visual marks that represent the data points). We will only be
doing a shallow exploration of what `ggplot2` can do. In this lab, we will typically 
define the data set using `ggplot()` and an aesthetic mapping using `aes()` and
then build on this object by adding layers.

## The basics and creating bar plots

When working with `ggplot2`, we can initialize plot objects and then layer on 
additional information to create our visualization. One of the simplest graphs
is a bar plot. This will be useful when analyzing a categorical variable or 
comparing counts. 

Let's import a data set. The `"presidential"` data set contains information about
the last 12 presidents of the United States. We can import it using `data("presidential")`. (This is a built-in data set for R, so you should not need to download anything.) 
Let's make a basic bar plot graphing the number of times each party help the presidency.
We can create a `ggplot` object like so
``` {r, include=TRUE}
data("presidential")
PresidentsBarPlot <- ggplot(data = presidential, aes(x = party))
```
This designates the data to use (the presidential data set) and the 
x-cooridinate (the party of the president), but we still need to specify how to 
visualize the data. We can do this by adding a `geom_bar()`. We can view the 
graphic by using `print()`.
```{r, include = TRUE}
print(PresidentsBarPlot + geom_bar(stat="count"))
```
Note that we did not need to specify the data and
coordinates again since this was saved in the `PresidentsBarPlot` object. Also, we added on additional features to the object by doing `+ <layer>`. This 
is getting us closer, but it's missing a title. Also, we'd like to color the 
bars to make it a little more interesting. Additional features can be added by
creating extra layers. 

``` {r, include = TRUE}
data("presidential")
PresidentsBarPlot <- ggplot(data = presidential, aes(x = party, fill=party)) +
  geom_bar(stat="count") +
  scale_fill_manual(values = c("blue", "red")) + 
  ggtitle("Parties of recent presidents")
print(PresidentsBarPlot) 
```

The above bar plot is looking better. We now have a title due to adding 
`ggtitle("Parties of recent presidents")`. We can designate the color of the 
bars based on the party with the additional argument `fill = party`. The line
`scale_fill_manual(values = c("blue", "red"))` changes the default colors to 
the more traditional political party colors.

**Task**: You will create your own box plot as above. Do the following to build 
your plot.

- Import the "mpg" data set using `data("mpg")`. This data set include details of 
cars manufactured from 1999 to 2008.
- Create a ggplot object called `CarPlot` where the x-coordinate and the fill color are given by `class`.
- Add a `geom_bar(stat="count")` and `ggtitle()` layer to `CarPlot` and print the result.
If you did everything correct, you should get a plot like below.

```{r, echo=FALSE}
data("mpg")
CarPlot <- ggplot(data = mpg, aes(x = class, fill=class)) + geom_bar(stat = "count")
print(CarPlot + ggtitle("Class of cars"))
```

## Estimating distributions with histograms and violin plots

Moving now to analyzing quantitative variables, we generally want graphs which 
will give us an idea about the distribution of the variable. One way to do this
is with a histogram. Creating one is similar to a bar plot.

```{r}
GasPlot <- ggplot(data = mpg, aes(x= hwy)) + 
  geom_histogram(fill="blue", alpha = 0.5, binwidth = 2.5) +
  ggtitle("Distribution of Highway Gas Mileage") +
  labs(x="Highway Mileage")
print(GasPlot)
```

Another popular option for getting a sense of the distribution of a variable is
with violin plots. We can get an idea for how to read such a chart with an example.

```{r}
data("iris")
IrisPlot <- ggplot(data = iris, aes(x=Species, y = Petal.Length, fill=Species)) + 
  geom_violin() +
  geom_dotplot(binaxis='y', stackdir='center', dotsize=0.5, fill="white") +
  ggtitle("Distribution of Petal Length by Species of Iris")
print(IrisPlot)
```
Here we have plotted a violin plot for each species of iris in order to determine
the distribution of petal lengths. The outside of the violin plot will expand if 
more points are located near there, making the plot look like a violin. Plotting
the points of the distribution on top of the graphs emphasizes this (although
we do not *need* the points for the plot). This is a convenient way to quickly compare the 
distributions of several populations quickly.

**Task**: Recreate the graph below as follows:

- Import the `iris` data set.
- Construct a violin plot of Petal Width by species. Make the fill color the
species type.
- Add a title and set the colors of the fill to "purple", "pink", and "lavender".

```{r, echo=FALSE}
data("iris")
IrisPlot2 <- ggplot(data = iris, aes(x=Species, y = Petal.Width, fill=Species)) + 
  geom_violin()  +
  ggtitle("Distribution of Petal Width by Species of Iris") +
  scale_fill_manual(values = c("purple", "pink", "lavender")) 
print(IrisPlot2)
```

## Scatterplots

Finally, if we want to compare two quantitative variables, then scatter plots will
be the most appropriate way to visualize this. Returning against to the iris data
set, let's see how sepal length is related to petal length

```{r}
data("iris")
IrisScatter <- ggplot(data = iris, aes(Sepal.Length, Petal.Length)) + 
  geom_point() + 
  ggtitle("Petal Length vs Sepal Length")
print(IrisScatter)
```

There seems to be an obvious correlation between these variables. But there are
clusters in the data. What is causing this? We can get some insight by coloring 
based on species.

```{r}
data("iris")
IrisScatter <- ggplot(data = iris, aes(Sepal.Length, Petal.Length, color=Species)) + 
  geom_point() + 
  ggtitle("Petal Length vs Sepal Length")
print(IrisScatter)
```

By coloring, we can see the clustering is caused by the different species. We can get
a better idea about the coorelations within the species by plotting lines of best
fit.

```{r}
print(IrisScatter + geom_smooth(method = lm))
```
**Task**: Recreate the plot below by doing the following:

- Import the data set `mpg` using `data("mpg")`. Then filter the data set to only
include cars with class "compact" or "suv". *Hint*: the code `mpg$class %in% c("compact", "suv")` will give you a vector of logical values correspond to the rows where
the class is either "compact" or "suv". You can use this to filter the data set.
- Create a scatter plot with the filtered data set. Set the color to correspond
with the class of car.
- Set the title using `ggtitle(...)` and the axis labels using `labs(x=...,y=...)`.


```{r, echo=FALSE}
compact_and_suv = mpg[mpg$class %in% c("compact", "suv"), ] 
CarScatter <- ggplot(data = compact_and_suv, aes(cty, hwy, color=class)) +
  geom_point() + 
  labs(x="City", y="Highway") +
  ggtitle("Highway vs City Gas Mileage")
print(CarScatter)
```

## Moving forward

You hopefully have some idea of what can be done using `ggplot2` after this lab.
I'd encourage you to try importing some of your own data into R and making some 
plots of your own. When working, make sure to save the code to construct your plots
so that you can recreate them if necessary. 